---
title: "An Introduction to data analysis in R"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

# Hello data analysis 

## Introduction 

In this tutorial we consider data introduced by Ronald Fisher in his 1936 paper *The use of multiple measurements in taxonomic problems*, otherwise known as *Iris*. 

The learning objective is to develop experience of data analysis using some R commands for summary statistics, plots, correlation and group comparisons.

## Load data

### Inbuilt data in R

The data we use in this tutorial contains three plant species (setosa, virginica, versicolor) and four features measured for each sample. These quantify the morphologic variation of the iris flower in its three species, all measurements given in centimeters.

This dataset is inbuilt into R. It can be accessed via:

```{r, echo=TRUE}
## load iris, a dataset inbuilt into R
data(iris)
```

The first thing I do when loading data is to check it looks as expected. The command `head()` shows the first few rows.

```{r, echo=TRUE}
## have a look at the data: head() prints out the first few rows
head(iris)
```

### Saving and loading your own data

When you have your own dataset you will need to load it into R in a different way. 

A common method is to save data as a commar separated file (.csv). This is a text file that separates out columns using a commar (,). You can do this in many programs, including Microsoft Excel, Libreoffice or just a text editor.

Data can be saved in R as a csv file using `write.csv()`. We next demonstrate how to save the data from R to a CSV file, in this case the iris data:

```{r, echo=TRUE}
write.csv(file="iris-out.csv", iris, row.names=FALSE)
```
This created a CSV file called `iris-out.csv`.  

How can we load a CSV file? One way is through `read.csv()`. For example, to load the iris data into an object we call *iris* use:

```{r, echo=TRUE}
iris = read.csv("iris-out.csv")
```

*Now your turn: write the command to display the first few rows on 'iris' data*
```{r head-ex, exercise=TRUE}
head(__) 
```
<div id="head-ex-hint">
**Hint:** replace __ with iris
</div>


## Summary statistics

### `summary()` function

In the first tutorial you were introduced to the `summary()` function. When run on a data frame like the `iris` data it gives summary statistics for each column, including min, max, mean, median and upper and lower quartiles:

```{r, echo=TRUE}
summary(iris)
```

### `apply()`: statistics by row or column of data

A more general way to get statistics by *column* is to use `apply(__,  2, __)`. For example, let us calculate the mean to the first 4 columns of iris 

```{r, echo=TRUE}
apply(iris[,1:4], 2, mean) #mean first 4 cols
```
In general we get statistics by *column* through `apply(*data*,  2, *statistic*)` where we tell R
 
- what *data* to use (e.g. `iris`), 
- and what *statistic* or function to run (e.g. `mean`),

The `2` says do it by *column*. If we wanted to for for each rows in the data we'd replace with `1`. 

So, we can do standard deviation by column by changing `mean` for `sd`, via:

```{r, echo=TRUE}
apply(iris[,1:4], 2, sd) #standard deviation
```
*Now your turn - calculate the maximum by column *

```{r applyex1, exercise=TRUE}
apply(iris[,1:4], 2, __) 
```

<div id="applyex1-hint">
**Hint:** replace __ by `max`
</div>

When presenting data it is important to remove spurious precision and many decimal places, by rounding. 
The `round()` function is for this. It is demonstrated on the standard deviation by column below.

```{r, echo=TRUE}
##save to R object called mysd
mysd = apply( iris[,1:4], 2, sd ) #standard deviation

## print out, rounded to 2 d.p.
round(mysd, 2)
```
The second argument to `round()` says how many decimal places to round to. Above I rounded to 2 d.p..

*Now your turn. In the code cell below, try to calculate standard quantiles with the R function `quantile`, and round them to 1 decimal place *

```{r roundapplyex, exercise=TRUE}
myquant = apply( iris[,1:4], 2, __ ) 

round(__, __)

```
<div id="roundapplyex-hint">
**Hint:** see the command used above - change `sd` to `quantile` and `mysd` to `myquant` 
</div>


### `tapply()`: statistics by group

`apply()` calculates a function by row or column. What if you want to do the summary statistic by different groups? For example, we might with to know summary statistics for each different iris species. One way is to do this is to use `tapply()`. The next code chunk calculates mean sepal length by species:


```{r, echo=TRUE}
tapply(iris$Sepal.Length, iris$Species, mean) #mean sepal length by type of plant
```

In the function `tapply(argument 1, argument 2, argument 3)`

- The *first* argument is the *data of interest* (iris\$Sepal.Length in example above), 
- The *second* argument is the *grouping variable* (here iris\$Species), and 
- The *third* argument is the function or *statistic* we want to calculate (here mean).

This is another example that does summary statistics of sepal length by species:

```{r, echo=TRUE}
tapply(iris$Sepal.Length, iris$Species, summary) 
```
*Now your turn. Calculate standard deviation of sepal length by species *

```{r tapplyex1, exercise=TRUE}
tapply(iris$Sepal.Length, iris$Species, __) 

```
<div id="tapplyex1-hint">
**Hint:** the statistic is `sd`.
</div>

*Now calculate summary statistics about petal width by species *

```{r tapplyex2, exercise=TRUE}
tapply(iris$__, iris$Species, __) 

```
<div id="tapplyex2-hint">
**Hint:** The column name for petal width is `Petal.Width`. The statistic is `summary`.
</div>



## Histograms

In the previous section the summary statistics showed that petal length median was much higher than the mean. This is illustrated by the overall distribution of petal length data shown using a histogram, by using the function `hist()` as follows.

```{r, echo=TRUE}
hist(iris$Petal.Length)
```

The median is much higher than the mean because there is a large spike close to zero in the histogram. The reason is we have mixed up three species in the overall distribution - setosa, versicolor and virgina. 

If we look within each species (or subgroup) the median is close to the mean. 

*You turn: use `tapply()` to calculate summary statistics by Species *

```{r extapsum, exercise=TRUE}
tapply(iris$Petal.Length, iris$__, summary) 
```

We may also wish to visualise histograms for each subtype. One way to do this is to use the `subset()` operator introduced earlier. First define the dataset (I call it `setosa`) that only includes this Species, then run the histogram for Petal length. 

```{r, echo=TRUE}
setosa = subset(iris, iris$Species == "setosa")

hist(setosa$Petal.Length)
```


*Now your turn. Produce a histogram of petal length for the versicolor type of iris. *

```{r extapsum2, exercise=TRUE}
versicol= subset(iris, iris$Species == __)

hist(__$Petal.Length)

```
<div id="extapsum2-hint">
**Hint:** The Species is called "versicolor". Don't forget the quotation marks. The name of the new data with just veriscolor types needs to be entered into the `hist()` command.
</div>

Sometimes it helps to put all charts on the same plot. This may be accomplished, for example, as follows.

```{r, echo=TRUE, fig.width=6, fig.height=12}
setosa = subset(iris, iris$Species == "setosa")

versicol = subset(iris, iris$Species == "versicolor")

virginica = subset(iris, iris$Species == "virginica")

par(mfrow=c(3,1))

hist(setosa$Petal.Length, xlim=c(1,7), col=1, main="setosa", breaks=seq(1,7,by=0.25), xlab="Petal length")

hist(versicol$Petal.Length, xlim=c(1,7), col=1, main="versicolor", breaks=seq(1,7,by=0.25), xlab="Petal length")

hist(virginica$Petal.Length, xlim=c(1,7), col=1, main="virginica", breaks=seq(1,7,by=0.25), xlab="Petal length")

```

The plots shows clear differences in petal length between the species. 

The histograms also look a little different than earlier on, because we did the following in the code.

- Before calling `hist()` we wrote `par(mfrow=c(3,1))`. This instructs R to have three plots arranged in the single plot, as a matrix with *3* rows and *1* column. The next commands fill up the slots. If you wanted them with three columns and one row then you'd use `par(mfrow=c(1,3))`; a 2x2 matrix would be `par(mfrow=c(2,2))` etc.

- We gave more arguments to `hist()` to customise the plots. These were:
  
- Choosing the x axis range (`xlim = c(1,7)`). Can similarly set y-axis through `ylim = c(__, __)`

- Choosing the title of the plot through `main="my title"`

- Choosing colour (bars as black, through `col = 1`)

- Choosing the bins through `breaks = `(and use of `seq()` function for a sequence, where `seq(1,7,by=0.25)` is shorthand for c(1,1.25, 1.5, 1.75, 2, 2.25, ..., 6.75, 7))

- Setting x-axis label through `xlab=""`, similarly y-axis label can be set   



*Now your turn. Produce two histogram of petal width using the `setosa` and `virginica` type of iris, restricting the x-axis betwee 0 and 2.5. Make them red by setting `col=2`. *


```{r extapsum3, exercise=TRUE,fig.width=6, fig.height=8} 
setosa = subset(iris, iris$Species == "setosa")

versicol = subset(iris, iris$Species == "versicolor")

par(mfrow=c(__,1))

hist(__$Petal.Width, xlim=c(0, __), col=__, breaks=seq(0, 2.6, by=0.2)) 

hist(__$Petal.Width, xlim=c(0, __), col=__, breaks=seq(0, 2.6, by=0.2))


```
<div id="extapsum3-hint">
**Hints:** We want two rows in the plot - enter 2 into the `mfrow` part. Then, inside `hist` you need the data names first, 2.5 in the `xlim` argument, and col=2. (If you change col=3 what happens?)
</div>



## Boxplots

An alternative way to show the distribution is through a boxplot.

In R if you call the `plot()` function with a *factor* variable (such as Species) first, followed by the numeric variable (such as petal length) then it will produce a boxplot. For example

```{r plotmea, exercise=TRUE}
plot(iris$Species,iris$Petal.Length, ylab="Petal Length")
```

An alternative is to use the function `boxplot()`. 

*Your turn: create a box plot of petal length for virginica subtype *
```{r boxplotex, exercise=TRUE}
virginica = subset(iris, iris$Species == "virginica")

boxplot(virginica$__)
```
<div id="boxplotex-hint">
**Hint:** Petal.Length
</div>

It is also possible to put boxplot of subgroups side by side using the `split()` function

```{r plotmeb, exercise=TRUE}
boxplot( split( iris$Petal.Length, iris$Species) )
```

*Task for you: adapt the code just introduced to produce a boxplot of petal width by species using the `split()` function. Label the y-axis "Petal Width". *

```{r boxplotex2, exercise=TRUE}
boxplot( __( iris$__, iris$Species), ylab=__ )
```

<div id="boxplotex2-hint">
**Hint:** Use `split`, Petal.Width. Don't forget quotation marks for the label. 
</div>



## Scatter plots

Next lets look at the association between variables by using scatter plots. First plot petal length against sepal length. 

```{r plotmec, exercise=TRUE}
plot(iris$Petal.Length, iris$Sepal.Length)
```

This combines the different species. To look at them separately one might plot them separately. 

*Task for you: repeat the plot with `setosa` and `virginica` data only *
```{r scatterex1, exercise=TRUE}
setosa = subset(iris, iris$Species == "setosa")

versicol = subset(iris, iris$Species == "versicolor")

par(mfrow=c(1,2))

plot(__$Petal.Length, __$Sepal.Length)

plot(__$Petal.Length, __$Sepal.Length)
```

A better aproach is to include all species on the same plot, but to use different plotting characters.

- To do this we'll first create a variable called *iris$colour* that is a number 1,2,3 related to the species. This is so that we can use different colour and plotting symbols in the charts.
- Then we use the `col` option to choose colour, and
- `pch` option to choose the plotting character 


```{r plotmed, exercise=TRUE}
iris$colour = as.integer( iris$Species ) ##Species is a factor, as.integer() converts it to a number to be used for colour choice


plot(iris$Petal.Length, iris$Sepal.Length, col = iris$colour, pch=iris$colour)

```

*Task for you: do the same plot but for Petal Width vs Petal Length, and add x-label and y-labels *
```{r scatterex2, exercise=TRUE}
iris$colour = as.integer( iris$Species ) ##Species is a factor, as.integer() converts it to a number to be used for colour choice


plot(iris$__, iris$__, col = iris$colour, pch=iris$colour, xlab=__, ylab=__)

```
<div id="scatterex2-hint">
**Hints:** Petal.Length; Petal.Width; Don't forget the quotation marks for the x and y labels; The first argument to plot is the x-axis.
</div>

R has a poor choice of colours by default, because some people cannot see the difference between red and green. But we can change the colours as we like. For example, suppose we want 1 to translate to black, 2 to purple and 3 to orange.

```{r plotmecol, exercise=TRUE}
iris$colour = as.integer( iris$Species ) ##Species is a factor, as.integer() converts it to a number to be used for colour choice

plot(iris$Petal.Length, iris$Sepal.Length, col = c("black", "purple", "orange")[iris$colour], pch=iris$colour)

```

It is possible to add a legend to the chart, to identify which points correspond to which species by using the `legend()` function.


```{r plotmee, exercise=TRUE}

iris$colour = as.integer( iris$Species ) ##Species is a factor, as.integer() converts it to a number to be used for colour choice

plot(iris$Petal.Length, iris$Sepal.Length, col = c("black", "purple", "orange")[iris$colour], pch=iris$colour)

legend("topleft", c("Setosa", "Versicolor", "Virginica"), col=c("black", "purple", "orange"), pch=1:3)
```

The first argument specifies where to place the key ("topleft"). The second are the labels to use; `col` the colours; `pch` the plotting characters. You need to make sure these are the same as specified in the plot.

*Task for you: do the same plot but give for Petal Width vs Petal Length, and add x-label and y-labels. Then add the legend to "bottomright", using half size characters (`cex`=0.5)*
```{r scatterex3, exercise=TRUE}

iris$colour = as.integer( iris$Species ) ##Species is a factor, as.integer() converts it to a number to be used for colour choice

plot(iris$__, iris$__, pch = iris$colour, col=c("black", "purple", "orange")[iris$colour], xlab=__, ylab=__)

legend(__, c("Setosa", "Versicolor", "Virginica"), col=c("black", "purple", "orange"), pch=1:3, cex=0.5)

```

Lastly, a quick way to plot all the variables (Petal and sepal length and width) against each other in the same chart is to use a function called `pairs()`. This does a so-called matrix plot of the variables against each other.

```{r plotshow, exercise=TRUE}
iris$colour = as.integer( iris$Species ) ##Species is a factor, as.integer() converts it to a number to be used for colour choice

pairs(iris[,1:4], pch=iris$colour, col=c("black", "purple", "orange")[iris$colour])
```

This shows correlation between the measurements, and also clear differences between species (setosa black o, versicolor purple triangle, virginica orange +).

## Correlation coefficients

R has functions to calculate correlation coefficients to quantify the association between two continuous variables. The main one is `cor.test`. For example, 

```{r, echo=TRUE}
cor.test(iris[,1], iris[,2])
```

The first column is sepal length, the second sepal width, so this is correlation between sepal length and width. 

An alternative way to run the same command is name the columns.

*Task for you: calculate Spearman correlation between Sepal Length and Sepal Width by naming these columns, and using the additional argument method="spearman"* 

```{r cor-check, exercise=TRUE}
cor.test( iris$__, iris$__, method="spearman")
```

*Task for you. Calculate spearman correlation for the setosa subgroup, using the setosa subset data *


```{r corexer, exercise=TRUE}
setosa = subset(iris, iris$Species == "setosa")

cor.test(__$Sepal.Length, __$Sepal.Width, method= __) 
```

There is a much stronger correlation when on species basis only, as expected.

*Q Now try to repeat the above analysis for a different species, and different measurements - you decide what*

```{r cor-exex, exercise=TRUE}
##add your code here
```

## Two-group comparisons

We next use the iris data to demonstrate some methods used to compare two independent groups.

We demonstrate by looking at the sepal length in the versicolor and virginica species (i.e. two groups).

Suppose we wish to understand differences between the species in Sepal length. We might first look at summary statistics and boxplots, as already introduced.

```{r, echo=TRUE}
summary(versicol$Sepal.Length)

summary(virginica$Sepal.Length)
```

```{r, echo=TRUE}
boxplot(versicol$Sepal.Length, virginica$Sepal.Length, names=c("Versicolor", "Virginica"))
```

The groups certainly seem different. To test for a difference in means one might do a t-test, using the `t.test()` function as follows.

```{r, echo=TRUE}
t.test(versicol$Sepal.Length, virginica$Sepal.Length)
```

The output does a t-test. It gives the t-statistic and associated p-value. It also estimates the mean difference between the groups with a 95% confidence interval for the mean difference. 

*How would you do a t-test for Sepal Width? *

```{r ex-ttest, exercise=TRUE}
versicol = subset(iris, iris$Species == "versicolor")

virginica = subset(iris, iris$Species == "virginica")

t.test(versicol$__, virginica$__)
```

The Wilcoxon (non-parametric) test follows the same structure, but is called `wilcox.test()`.

*Try to do a wilcoxon test for Sepal Length *

```{r ex-wilctest, exercise=TRUE}

versicol = subset(iris, iris$Species == "versicolor")

virginica = subset(iris, iris$Species == "virginica")

__(versicol$Sepal.Length, virginica$Sepal.Length)

```

## Linear regression

We finally use the iris data to demonstrate linear regression in R.

We demonstrate by looking at the sepal length vs petal length, firstly for the versicolor species.

We first plot the data to assess the relationship visually.

```{r, echo=TRUE}
plot(versicol$Petal.Length,versicol$Sepal.Length, xlab="Sepal Length (cm)", ylab="Petal length (cm)")
```

As expected, there does appear to be an association. Next we fit a linear regression model to these data, and print the summary and confidence intervals on the regression coefficients.

```{r, echo=TRUE}
myreg<-lm(Sepal.Length ~ Petal.Length, versicol)
summary(myreg)
round(confint(myreg),2)
```
```{r question-absrisk3}
question_radio(
  "What is the estimated intercept and 95%CI to 2 decimal places (dp))?",
  answer("2.41 (95%CI 1.51 to 3.30)", correct=TRUE),
  answer("0.82 (95%CI 0.62 to 1.04)"),
  answer("2.41 (95%CI 0.45 to 5.40)"),
  answer("0.83 (95%CI 0.10 to 7.95)"),
  allow_retry = TRUE,
  random_answer_order=TRUE)
```

We may add the fitted regression line to the scatter plot as follows.

```{r, echo=TRUE}
##scatter plot as above
plot(versicol$Petal.Length,versicol$Sepal.Length, xlab="Sepal Length (cm)", ylab="Petal length (cm)")
## add grid based on x/y axis ticks
grid()
## add regression line
abline(myreg, col="red")
# add confidence intervals to regression line
## x-values at which to calculate confidence interval
newx <- seq(2, 6, by=0.2)
## confidence interval at the x-values
conf_interval <- predict(myreg, newdata=data.frame(Petal.Length=newx), interval="confidence",level = 0.95)
## add the confidence interval lines to plot
lines(newx, conf_interval[,2], col="blue", lty=2)
lines(newx, conf_interval[,3], col="blue", lty=2)

```


*Based on the fit shown in the plot, do you think a line looks a good model for the data?*

It looks reasonable to my eye. To illustrate what could happen when a linear model is not a good fit, and how the plots above look, we next repeat the analysis but fit to all species together.


```{r, echo=TRUE}
plot(iris$Petal.Length,iris$Sepal.Length, xlab="Sepal Length (cm)", ylab="Petal length (cm)")
```

Just by ploting the data we can see that a linear might be reasonable for the right hand side (from approx 3 on the x-axis), but not to the whole data.

*Q: Repeat the regression analysis above, but using the complete data which is in the `iris` dataframe. Replace __ below with the correct text to do this.*.


```{r repeat-ref2, echo=TRUE, exercise=TRUE}
##scatter plot as above
plot(__$Petal.Length, __$Sepal.Length, xlab="Sepal Length (cm)", ylab="Petal length (cm)")
## add grid based on x/y axis ticks
grid()
## fit regression
myreg2<-lm(Sepal.Length ~ Petal.Length, iris )
## add regression line
abline(__, col="red")
# add confidence intervals to regression line
## x-values at which to calculate confidence interval
newx <- seq(1, 8, by=0.2)
## confidence interval at the x-values
conf_interval <- predict(__, newdata=data.frame(Petal.Length=newx), interval="confidence",level = 0.95)
## add the confidence interval lines to plot
lines(newx, conf_interval[,2], col="blue", lty=2)
lines(newx, conf_interval[,3], col="blue", lty=2)

```
<div id="repeat-ref2-hint">
**Hint:** replace __ with iris, or myreg2 as appropriate
</div>




## Conclusion

This is the end of the tutorial. A number of R functions for basic data analysis have been introduced. 

The next tutorial will introduce you to RStudio. Once you have done this, you are suggested to try to run the analysis from this tutorial in RStudio by copying the commands in order.
